Leadership across cultures and contexts: an exploratory analysis

### 2020-06-22 leadership_across_cultures_contexts.Rmd
Changing format. Fixing citation
More edits to reduce words

These four factors were confounded. Evidence for leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013; @venkataraman_hunter-gatherer_2017]. Evidence for supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories (Figures \@ref(fig:grouptype) and S\@ref(fig:regionsubsistence)). These patterns support perspectives suggesting residential mobility and subsistence intensification are associated with increasing complexity of sociopolitical organization [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014]. One interpretation of this trend is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes.

@@

Several theories of leadership also emphasize the moral authority of leaders [e.g., @henrich_big_2015; @henrich_evolution_2001; @richerson_tribal_2009; @tietjen_moral_1985; @boggild_evolved_2016], yet evidence for leaders as sources of moral authority or "fair" was found in only a minority of cultures (`r moral_pct`% and `r fairness_pct`% of cultures, respectively). The dominance theory of leadership [e.g., @chapais_competence_2015; @cheng_dominance_2019; @cheng_pride_2010; @henrich_evolution_2001; @tiger_imperial_1971] emphasizes leaders' ability to instill fear, but in only `r feared_pct`% of cultures was there evidence that leaders were feared. Finally, a large literature on leadership in industrialized societies highlights the physical attractiveness of leaders [e.g., @sinclair_sexuality_1995; @altemeyer_sexual_1974; @march_leadership_2009], but evidence that leaders were attractive was found in only `r attractive_pct`% of the nonindustrial cultures sampled here. Clearly there is substantial diversity in many dimensions of leadership among nonindustrial populations.

@@

The evolution of leadership and followership is somewhat perplexing. Leadership often involves the provisioning of a public good at an individual cost, raising the question of how the individual benefits of leadership outweigh the costs. Followership, on the other hand, involves relinquishing individual decision-making to a leader who might or might not act in the follower's interest. The benefits of followership (and by extension, group living) must also outweigh the costs for leadership to evolve and be maintained [e.g., @gavrilets_solution_2014; @hooper_theory_2010; @price_evolution_2014; @bastardoz_nature_2018; @hagen_leadership_2019; @garfield_review_2019].

@@

Results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Followers were not reported to receive mating benefits, relative to leaders or other benefit types.

@@

Relationships between leadership and high status are currently the focus of much work in evolutionary leadership studies [e.g., @van_vugt_dual_2019; @smith_leadership_2020; @cheng_dominance_2019; @von_rueden_dynamics_2019]. Although leadership and high social status often covary, leaders are not always high-status individuals, nor are high status individuals always leaders [see, e.g., @van_vugt_dual_2019; @cheng_why_2020; @garfield_review_2019].

@@

In our view, because individuals maintain multiple social "statuses," social context is critical [see @wiessner_power_2010]. In many cultures, for example, communities form work-groups and appoint "chiefs-for-a-day" who manage the work but have little status or authority otherwise [e.g., @macfarlan_competitive_2012]. @garfield_evolutionary_2019 found that leader organization dimensions clustered separately from leader prestige dimensions, suggesting a distinction. @garfield_investigating_2020 found that some leaders among recently settled hunter-gatherers had organizing responsibilities but unexceptional levels of respect. Some older individuals, on the other hand, were highly respected but had little social influence or authority.

@@

The words and dimensions of leadership predictive of high status leaders (Figure \@ref(fig:highstatusplot)) suggest the phenotype of such leaders resembles what many anthropologists refer to as "Big men" [@garfield_review_2019]: wealthy, accomplished, older men who favor members of their kin group, and are highly regarded for their decision-making abilities, knowledge, and intelligence. High status leaders are able to direct benefits to their close relatives, likely due, in part, to their greater material and social wealth, which implicates leadership dynamics as a driver of the evolution and emergence of inequality [c.f., @sanday_female_1981; @shenk_status_2016; @price_social_2010; @borgerhoff_mulder_pastoralism_2010; @mattison_evolution_2016; @bowles_emergence_2010]. Our results did not strongly implicate high status leaders with particular functions, with the exception that punishment was modestly negatively associated with evidence for high status. The word "economic" was a negative predictor of high status, hinting that organizational roles might not be high status.

@smith_leadership_2020 speculate that, "leadership may...be more strongly correlated with high status in large, complex organizations, such as in corporations and governments (p. 2)." Evidence for high status leaders, however, did not substantially vary by group structure, which included levels of social organization ranging from sub-residential to supracommunity (Figure S\@ref(fig:groupstatus)). This null result is evidence against the @smith_leadership_2020 hypothesis.

@@

<!-- PROSCIAL FUNCTIONS FEATURE -->
<!-- DISCUSSES PROSOCIALITY FEATURE -->
**Prosociality**. The prosocial cluster (Figure \@ref(fig:cluster)A), comprising prosocial competencies such as generosity, interpersonal and oratory skills, and charisma, and cultural conformity, such as proper behavior and having a favorable personality, generally corresponds to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @grabo_charisma_2017; @howell_role_2005]. The importance of leader prosociality also supports the @henrich_big_2015 Big-man mechanism by which positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership.
<!-- A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity). -->

We also found leader functions often involve prosocial investments in group welfare (Figure \@ref(fig:cluster)B). The prosociality feature of leader functions included distributing resources, providing protection, and various prosocial investments, which are functions commonly highlighted in the anthropological literature on leadership in nonindustrial societies [@sahlins_poor_1963; @johnson_organizational_1982; @service_origins_1975; @boehm_egalitarian_1993; @boehm_hierarchy_1999; @johnson_evolution_1987]. In the absence of institutionalized leadership roles, leaders often maintain social influence via continued community investment and the demands of followers can shape leadership functions to meet local needs.

@@

Leaders, possibly due to their special competencies, are able to devise and implement strategic solutions to group problems including resolving conflicts within the group (the most widely supported function in our sample) and providing counsel (also widely supported), which again are functions that are commonly noted in the anthropological and psychological leadership literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991; @de_cremer_intergroup_2002]. The feature of leader organizational functions aligns with many evolutionary theories that emphasize the role of leaders in organizing collective actions [e.g., @hooper_theory_2010; @van_vugt_cognitive_2007; @glowacki_formation_2016; @pietraszewski_evolution_2020; @perret2020].

@@

In our view, the dominance-prestige and "dual-model" theories of leadership, while advancing the discipline, cannot easily account for shaman leaders nor "chiefs-for-a-day" (managers). Critically, these models have conflated the ability to provide "benefits" with knowledge and expertise, and the ability to impose "costs" with physical formidability. @van_vugt_dual_2019, for instance, state that the source of deference for prestige-style leadership is "Information asymmetry", and for dominance-style, "Power asymmetry" (their Table 1). Instead, as shaman leaders demonstrate, knowledge and expertise can both provide benefits and impose costs, and the same goes for physical formidability. The use of knowledge to impose costs is widely recognized by many social scientists. The role of gossip and other forms of indirect aggression in social dominance and achieving leadership roles, for instance, has been extensively investigated in industrial populations [e.g., @hawley_ontogenesis_1999; @hess_gossip_2017; @hess_sex_2006] and also among hunter-gatherers [@hess_interpersonal_2010].

Our revised model, inspired by @von_rueden_roots_2014 and others, proposes that leaders deploy cognitive, social, material, and somatic capital to provide benefits and impose costs on followers. If their abilities to provide benefits or impose costs are similar to other groups members, as in "chiefs-for-a-day", they acquire little status, but if their abilities are exceptional relative to other group members, they gain status via biological market mechanisms [@hammerstein_biological_2016; @garfield_evolutionary_2019; @hagen_leadership_2019; @pietraszewski_evolution_2020]. There was particularly strong evidence for cognitive and social capital, with cognitive capital associated with both providing benefits and imposing costs, particularly among shaman leaders (Figure \@ref(fig:fivefold)).

@@

This is an exploratory study, which means that our analyses were driven by the data rather than by *a priori* hypotheses. This has the advantage that unexpected patterns can be discovered, but the disadvantage that some random variation in the data can, and almost certainly will, be misinterpreted as meaningful patterns. In particular, our cluster analysis of leader qualities (Figure \@ref(fig:cluster)A) was sensitive to the choice of different clustering algorithms and inclusion or exclusion of dimensions with little evidence.

# Concluding remarks: Beyond the non-WEIRD 
<!-- CAN'T JUST BE NONWEIRD -->
In recognizing the limited generalizability of most extant data [@henrich_weirdest_2010], many social scientists are now diversifying their samples to better understand *human* psychology, behavior, organizations, and culture. We caution that the WEIRD people problem in behavioral science cannot be remedied simply by a "non-WEIRD" solution. Our results, and much of anthropology and ethnography, clearly demonstrate enormous diversity among populations that are now often categorized as "non-WEIRD", "traditional", or "small-scale." Anthropologists have often made the opposite mistake in essentializing diverse "others" and failing to recognize deep similarities among peoples of all cultures [@brown_human_1991; @said1979; @reyes-garcia_small-scale_2017].

@@

@@

<!-- CLOSING STATEMENT -->
Leaders across cultures rely a range of individual capacities, including cognitive, supernatural, material, and physical endowments, as well as their social networks, to organize group members, implement strategic actions, provide prosocial services to the group, and impose costs, all while conforming to cultural norms. Currently, no single theoretical perspective has yet captured the ethnographic reality of human leadership dynamics.

>>>>

These four factors were confounded. Evidence for leadership in residential subgroups and kin groups was relatively more common in hunter-gatherer and pastoralist societies, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013; @venkataraman_hunter-gatherer_2017]. Evidence for supracommunity political leadership was more common in horticultural and agricultural societies, which tend to rely heavily on cultivated land and maintain fixed, defensible territories (Figures \@ref(fig:grouptype) and S\@ref(fig:regionsubsistence)). These patterns support perspectives suggesting residential mobility and subsistence intensification are associated with increasing complexity of sociopolitical organization [@johnson_evolution_1987; @steward_basin-plateau_1938; @powers_evolutionary_2014]. Perhaps as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes.

@@

Several theories of leadership also emphasize the moral authority of leaders [e.g., @henrich_big_2015; @henrich_evolution_2001; @richerson_tribal_2009; @tietjen_moral_1985; @boggild_evolved_2016], yet evidence for leaders as sources of moral authority or "fair" was found in only a minority of cultures (`r moral_pct`% and `r fairness_pct`% of cultures, respectively). The dominance theory of leadership [e.g., @chapais_competence_2015; @cheng_dominance_2019; @cheng_pride_2010; @henrich_evolution_2001; @tiger_imperial_1971] emphasizes leaders' ability to instill fear, but in only `r feared_pct`% of cultures was there evidence that leaders were feared. Finally, a large literature on leadership in industrialized societies highlights the physical attractiveness of leaders [e.g., @sinclair_sexuality_1995; @altemeyer_sexual_1974; @march_leadership_2009], but evidence that leaders were attractive was found in only `r attractive_pct`% of the nonindustrial cultures sampled here.

@@

Leadership often involves the provisioning of a public good at an individual cost, raising the question of how the individual benefits of leadership outweigh the costs. Followership, on the other hand, involves relinquishing individual decision-making to a leader who might or might not act in the follower's interest. The benefits of followership (and by extension, group living) must also outweigh the costs for leadership to evolve and be maintained [e.g., @gavrilets_solution_2014; @hooper_theory_2010; @price_evolution_2014; @bastardoz_nature_2018; @hagen_leadership_2019; @garfield_review_2019].

@@

Followers similarly benefited from increased material and social resources, but also often lost material and social resources. Followers were not reported to receive mating benefits, relative to leaders or other benefit types.

@@

@@

Although leadership and high social status often covary, leaders are not always high-status individuals, nor are high status individuals always leaders [see, e.g., @van_vugt_dual_2019; @cheng_why_2020; @von_rueden_dynamics_2019; @garfield_review_2019]. In our view, because individuals maintain multiple social "statuses," social context is critical [see @wiessner_power_2010]. In many cultures, for example, communities form work-groups and appoint "chiefs-for-a-day" who manage the work but have little status or authority otherwise [e.g., @macfarlan_competitive_2012]. @garfield_evolutionary_2019 found that leader organization dimensions clustered separately from leader prestige dimensions, suggesting a distinction. @garfield_investigating_2020 found that some leaders among recently settled hunter-gatherers had organizing responsibilities but unexceptional levels of respect. Some older individuals, on the other hand, were highly respected but had little influence or authority.

@@

The words and dimensions of leadership predictive of high status leaders (Figure \@ref(fig:highstatusplot)) resemble "Big men" [@garfield_review_2019]: wealthy, accomplished, older men who are highly regarded for their decision-making abilities, knowledge, and intelligence, and who favor their kin. These associations implicate leadership in the evolution and emergence of inequality [c.f., @sanday_female_1981; @shenk_status_2016; @price_social_2010; @borgerhoff_mulder_pastoralism_2010; @mattison_evolution_2016; @bowles_emergence_2010]. Our results did not strongly implicate high status leaders with particular functions, with the exception that punishment was modestly negatively associated with evidence for high status. The word "economic" was a negative predictor of high status, hinting that organizational roles might not be high status.

@smith_leadership_2020 speculate that, "leadership may...be more strongly correlated with high status in large, complex organizations, such as in corporations and governments (p. 2)." Evidence for high status leaders, however, did not substantially vary by group structure, which included levels of social organization ranging from sub-residential to supracommunity (Figure S\@ref(fig:groupstatus)). This null result fails to support this hypothesis.

@@

<!-- PROSCIAL FUNCTIONS FEATURE -->
<!-- DISCUSSES PROSOCIALITY FEATURE -->
**Prosociality**. The prosocial cluster (Figure \@ref(fig:cluster)A), comprising prosocial competencies such as generosity and charisma, along with cultural conformity, generally corresponds to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and evolutionary psychology [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @grabo_charisma_2017; @howell_role_2005]. The importance of leader prosociality also supports the @henrich_big_2015 Big-man mechanism by which positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership.
<!-- A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity). -->

We also found leader functions often involve prosocial investments in group welfare (Figure \@ref(fig:cluster)B). The prosociality feature of leader functions included distributing resources and providing protection, which are commonly highlighted in the anthropological leadership literature [@sahlins_poor_1963; @johnson_organizational_1982; @service_origins_1975; @boehm_egalitarian_1993; @boehm_hierarchy_1999; @johnson_evolution_1987]. In the absence of institutionalized leadership roles, leaders often maintain social influence via continued community investment and the demands of followers can shape leadership functions to meet local needs.

@@

Leaders, possibly drawing on their special competencies, devise and implement strategic solutions to group problems including resolving conflicts (the most widely supported function in our sample) and providing counsel (also widely supported), as commonly noted in the literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991; @de_cremer_intergroup_2002]. The *Organize* feature aligns with many evolutionary theories that emphasize the role of leaders in organizing collective actions [e.g., @hooper_theory_2010; @van_vugt_cognitive_2007; @glowacki_formation_2016; @pietraszewski_evolution_2020; @perret2020].

@@

In our view, the dominance-prestige and "dual-model" theories of leadership, while advancing the discipline, cannot easily account for shaman leaders nor "chiefs-for-a-day" (managers). Critically, these models have conflated the ability to provide "benefits" with knowledge and expertise, and the ability to impose "costs" with physical formidability. @van_vugt_dual_2019, for instance, state that the source of deference for prestige-style leadership is "Information asymmetry", and for dominance-style, "Power asymmetry" (their Table 1). Instead, as shaman leaders demonstrate, knowledge and expertise can both provide benefits and impose costs, and the same goes for physical formidability. The use of knowledge to impose costs is widely recognized by many social scientists. The role of gossip and other forms of indirect aggression in social dominance and achieving leadership roles, for instance, has been extensively investigated in industrial populations [e.g., @hawley_ontogenesis_1999; @hess_gossip_2017; @hess_sex_2006] and also among hunter-gatherers [@hess_interpersonal_2010]. 
Our revised model, inspired by @von_rueden_roots_2014 and others, proposes that leaders deploy cognitive, social, material, and somatic capital to provide benefits and impose costs on followers. If their abilities to provide benefits or impose costs are similar to other groups members, as in "chiefs-for-a-day", they acquire little status, but if their abilities are exceptional relative to other group members, they gain status via biological market mechanisms [@hammerstein_biological_2016; @garfield_evolutionary_2019; @hagen_leadership_2019; @pietraszewski_evolution_2020]. There was particularly strong evidence for cognitive and social capital, with cognitive capital associated with both providing benefits and imposing costs, particularly among shaman leaders (Figure \@ref(fig:fivefold)).

@@

This is an exploratory study, which means that our analyses were driven by the data rather than by *a priori* hypotheses. This has the advantage that unexpected patterns can be discovered, such as the role of shaman leaders, but the disadvantage that some random variation in the data can, and almost certainly will, be misinterpreted as meaningful patterns. In particular, our cluster analysis of leader qualities (Figure \@ref(fig:cluster)A) was sensitive to the choice of different clustering algorithms and inclusion or exclusion of dimensions with little evidence.

# Concluding remarks: Beyond the non-WEIRD 
<!-- CAN'T JUST BE NONWEIRD -->
The WEIRD people problem in behavioral science cannot be remedied simply by a "non-WEIRD" solution. Our results, and much of anthropology, clearly demonstrate enormous diversity among populations that are now often categorized as "non-WEIRD", "traditional", or "small-scale." Anthropologists have often made the opposite mistake in essentializing diverse "others" and failing to recognize deep similarities among peoples of all cultures [@brown_human_1991; @said1979; @reyes-garcia_small-scale_2017].

@@

Leaders across cultures rely a range of individual capacities, including cognitive, supernatural, material, social, and physical endowments, to organize group members, implement strategic actions, provide prosocial services to the group, and impose costs, all while conforming to cultural norms. Currently, no single theoretical perspective has yet captured the ethnographic reality of human leadership.

@@

<!-- CLOSING STATEMENT -->
